---
title: "HespTuMV_modelNYSfull"
author: "emlombardi"
date: "2025-06-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Introduction
This script takes the raster stack generated and cleaned in Rmd file `HespTuMV_modelNAfull` and clips it to New York State. Then I run spatial models using background data (same as other script, clipped to NYS), presence and absence data for TuMV in H. matronalis population surveys across New York habitats. 

# Libraries and prep
```{r}
# Libraries & Prep
```{r libraries}
library(ggplot2)
library(wesanderson)
library(RColorBrewer)
library(dplyr)
library(terra)
library(sf)
library(raster)
library(geodata)
library(enmSdmX)
library(AOI)
library(mapview)
library(brglm2)
library(rnaturalearth)
library(rnaturalearthdata)
library(rgbif)
library(rJava)
library(pROC)

#set wd
setwd("/Users/elizabethlombardi/Dropbox (Old)/Research/Human Preferences Paper (Hesperis)/Hesperis TuMV SDM project")

#set AOI polygons
#define the Areas Of Interest 
ny = aoi_get(state = "NY")
ny.sf = st_as_sf(ny, crs = 4326)
```


## Read Rasters
You might already have them loaded, but if not, compile the full version (North America) from here, and clip. 

```{r}
#Read
raster.files <- list.files(path = "/Users/elizabethlombardi/Dropbox (Old)/Research/Human Preferences Paper (Hesperis)/Hesperis TuMV SDM project/Rasters/Input Rasters/", pattern = "\\.tif$", full.names = TRUE)

rasters <- lapply(raster.files, rast) #read them individually

ref <- rasters[[1]] #bioclim rasters chosen as the reference raster to which I will align other rasters

# Resample all others to match the reference
rasters_aligned <- lapply(rasters, function(r) {
  resample(r, ref, method = "bilinear")  # use "near" for categorical data
})

rastStack <- rast(rasters_aligned) 
names(rastStack) #later, the model function needs columns to be free of special characters

#rename the rasters
names(rastStack) <- gsub("[^a-zA-Z0-9]", "", names(rastStack))
names(rastStack)[1:19] <- paste0("bio", 1:19)
names(rastStack)[21] <- "footprint"
names(rastStack)[22] <- "humanpop"
names(rastStack) #good

```

## Read occurrence data
Load the population surveys from ELISA/grad school, then background data from the other script (clip to NYS)

```{r}
#################### NYS surveys
#load NYS survey data of TuMV in hesperis populations
#all data (individual samples from 2016)
tumv.nys <- read.csv("~/Dropbox (Old)/Research/Hesperis/April 2021 ELISA/inferred.tumv.AllYears.binary.csv")
tumv.nys.sf <- st_as_sf(tumv.nys, coords = c("long", "lat"), crs = 4326)

#concatenated by population
popn.tumv.nys <- read.csv("~/Dropbox (Old)/Research/Hesperis/April 2021 ELISA/popns.AllYears.binary.csv")
popn.nys.sf <- st_as_sf(popn.tumv.nys, coords = c("long", "lat"), crs = 4326)

###################### BACKGROUND
#load cleaned background locations (pseudo absence) from other script
bg_pnts_buff <- read.csv("~/Dropbox (Old)/Research/Human Preferences Paper (Hesperis)/Hesperis TuMV SDM project/host_backgroundpnts_hostBuffered.csv")

bg_pnts_buff_nys <- bg_pnts_buff %>%
  st_as_sf(coords=c("decimalLongitude", "decimalLatitude"), crs=4326) %>%
  st_filter(ny.sf, .predicate = st_within)

#grab environmental data for background locations
bg_clim_nys <- terra::extract(rastStack, bg_pnts_buff_nys, ID=FALSE) #this is the host-weighted background bioclim df
dim(bg_clim_nys) #163 sites
bg_clim_nys <- na.omit(bg_clim_nys) #158 sites

head(bg_clim_nys)
```
### create occurrence, bg and env dataframe
Wait I actually don't want to do this. I have true absence locations in the dataset. Not sure a background layer is helpful at all. Mabye just figure out what the environmental correlations are and build models that way?
```{r}
presbg <- c(rep(1, nrow(tumv.nys.sf)), 
             rep(0, nrow(bg_clim_nys)))
```

### predictor correlation
```{r}
library(corrplot)
env.mat <- cor(bg_clim_nys, method="pearson") #pearson correlation computation
print(env.mat)
corrplot(env.mat, method="color", type="lower", )

library(caret)
high_corr <- findCorrelation(env.mat, cutoff = 0.7) #identify highly correlated variables
nysStack <- rastStack[[-high_corr]] #removes, but perhaps not the 'right' variables
names(nysStack)

library(caret)
control <- rfeControl(functions = rfFuncs, method = "cv", number = 5)
results <- rfe(x = nysStack, y = response, sizes = c(1:10), rfeControl = control)

```

# Clip data and rasters to NYS
This is the step that clips background data and the 



