---
title: "HespTuMV_SDM_NorthAmerica"
author: "emlombardi"
date: "2025-04-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#may also need to set Java installation path
#Sys.setenv(JAVA_HOME = "/Library/Internet Plug-Ins/JavaAppletPlugin.plugin/Contents/Home+ /Library/Internet Plug-Ins/JavaAppletPlugin.plugin/Contents/Home")
```

# Intro
This script is the version to be published with the paper. It was built interatively though many previous scripts, and is especially dependent on Kass_HespTuMVplay.R, which is a script guided by the publication by Kass et al. on species distribution modelling best practices. 

The objective of this script is to identify environmental variables that predict TuMV presence in Hesperis matronalis populations across North America. 

# Libraries & Prep
```{r libraries}
library(ggplot2)
library(wesanderson)
library(RColorBrewer)
library(dplyr)
library(terra)
library(sf)
library(raster)
library(geodata)
library(enmSdmX)
library(AOI)
library(mapview)
library(brglm2)
library(rnaturalearth)
library(rnaturalearthdata)
library(rgbif)
library(rJava)
library(pROC)

#set wd
setwd("/Users/elizabethlombardi/Dropbox (Old)/Research/Human Preferences Paper (Hesperis)/Hesperis TuMV SDM project")

#set AOI polygons
#define the Areas Of Interest 
ny = aoi_get(state = "NY")
ny.sf = st_as_sf(ny, crs = 4326)

#just use a bounding box for North America 
bb.NA <- ext(-170, -30, 5, 85) # minx, maxx, miny, maxy
NA.aoi <- as.polygons(bb.NA, crs = "EPSG:4326")
plot(NA.aoi)

```

# Read Rasters
The environmental predictor variables were already selected, projected, aligned, etc. and then saved to my local computer. A separate script will be included for replication of this step. 

```{r}
#Read
raster.files <- list.files(path = "/Users/elizabethlombardi/Dropbox (Old)/Research/Human Preferences Paper (Hesperis)/Hesperis TuMV SDM project/Rasters/Input Rasters/", pattern = "\\.tif$", full.names = TRUE)

rasters <- lapply(raster.files, rast) #read them individually

ref <- rasters[[1]] #bioclim rasters chosen as the reference raster to which I will align other rasters

# Resample all others to match the reference
rasters_aligned <- lapply(rasters, function(r) {
  resample(r, ref, method = "bilinear")  # use "near" for categorical data
})

rastStack <- rast(rasters_aligned) 
names(rastStack) #later, the model function needs columns to be free of special characters

#rename the rasters
names(rastStack) <- gsub("[^a-zA-Z0-9]", "", names(rastStack))
names(rastStack)[1:19] <- paste0("bio", 1:19)
names(rastStack)[21] <- "footprint"
names(rastStack)[22] <- "humanpop"

names(rastStack) #good

```

# Load TuMV data
These are the occurrence locations (presence-only) for TuMV in Hesperis matronalis. Based on image analysis of community collected data from iNaturalist. 

I will also remove duplicates, clean, etc. 

```{r}
tumv.na <- read.csv("~/Desktop/Research/Hesperis/Hesperis Computer Vision project/TuMV presence coords 2022 CLEAN.csv")

#remove duplicate IDs
tumv.cl <- tumv.na %>%
  group_by(ID) %>%
  arrange(desc(notes != "")) %>%  # Prioritize rows where 'notes' is not empty
  slice(1) %>%  # Keep the first row per ID
  ungroup()

#convert to spatial vector
tumv.cl <- tumv.cl %>% 
  mutate(across(where(is.character), ~ suppressWarnings(as.numeric(.)))) %>%
 rename("decimalLongitude"=longitude) %>%
  rename("decimalLatitude"=latitude) %>%
  mutate(species = "TuMV")

tumv.cl <- tumv.cl %>%
  dplyr::filter(accuracy.meters < 1000) #207 occs with accuracy below 1000 meters

#tumv.check <- tumv.cl %>% dplyr::filter(accuracy.meters > 1000) #0 occs with accuracy worse than 1000 meters

occs_vect <- terra::vect(tumv.cl, geom=c("decimalLongitude", "decimalLatitude"), 
                         crs = enmSdmX::getCRS("WGS84"))

hist(occs_vect$accuracy.meters, breaks=50)

```

# Sampling Bias analysis
This chunk of code will create an estimate of sampling bias in the TuMV dataset. I'll use the R package `sampbias` to visually and quantitatively assess sources of biase in the spatial distribution of occurrences in the data. 

Note that this is a bit tricky for our analysis because the virus distribution is nested in the host distribution, which is impacted by humans. This is a true factor, not only a sampling bias. Furthermore, using iNat observations means that there is implictly an association between human density and TuMV-Hesperis interaction. That is okay, though; it's observational in this study, and the fact that host-human-virus spatial correlation is high is part of what I want to say. 

```{r}
occs_sampbias <- tumv.cl[, c("species", "decimalLongitude", "decimalLatitude")]
#restrict_sample <- terra::convHull(occs_vect) |> terra::buffer(width = 200 * 1000) |> sf::st_as_sf() #this is a convext hull, but very big
NAhull <- terra::convHull(occs_vect) |> terra::buffer(width = 200 * 1000) |> sf::st_as_sf() #subset to first 10 records
rast.pnw <- terra::crop(rastStack[[1]], NAhull)
#mask.pnw <- terra::mask(envs_terra$wc2.1_30s_bio_1, restrict_sample) #most computationally costly, but maybe more accurate? check

# Raster of study area
inp_raster <- terra::rast(rast.pnw[[1]], vals = 0)

#calculate the bias
bias <- sampbias::calculate_bias(occs_sampbias,
                                 res=terra::res(inp_raster)[1], mcmc_iterations = 10000,
                                 buffer=.05, 
                                 restrict_sample = NAhull)
summary(bias)
t<-plot(bias) #clearly we've got some biases. Specifically with roads. Unsurprising, really. 


#map the bias for funsies
biasStack <- sampbias::project_bias(bias)
raster::plot(biasStack, nr = 3,
             legend.mar = 5.7)
mapview(biasStack$Total_percentage)

#project the biases to same projection as records
biasStack <- terra::project(biasStack, "epsg:4326")
names(biasStack)

```

### Load Bias Rasters
If you just want to reload the bias rasters, do so like this:

```{r}
####READING BIAS RASTERS FROM LOCAL COMPUTER
bias.folder <- list.files(path = "/Users/elizabethlombardi/Dropbox (Old)/Research/Human Preferences Paper (Hesperis)/Hesperis TuMV SDM project/Rasters/Bias Rasters/", pattern = "\\.tif$", full.names = TRUE)

bias.rasts <- lapply(bias.folder, rast) #read them individually

ref <- rasters[[1]] #bioclim rasters chosen?

# Resample all others to match the reference
rasters_aligned2 <- lapply(bias.rasts, function(r) {
  resample(r, ref, method = "bilinear")  # use "near" for categorical data
})

biasStack<- rast(rasters_aligned2)
names(biasStack)

#use the Total_percentage raster to 'down weight' areas around all sources of bias
samp_weight_rast <- biasStack[["Total_percentage"]]
bias_rel_max <- terra::extract(samp_weight_rast, occs_vect) #I think this is where the 2 NA values are introduced. 

#rescale 
bias_rel_max <- bias_rel_max[, 2]
occ_weights <- abs(bias_rel_max) / 100

```

### Plot Bias
Quick check of sources of bias across North America, clipped to TuMV presence hull

```{r}
# Convert raster to a dataframe for ggplot
bias.total <- biasStack$Total_percentage
bias_df <- as.data.frame(biasStack, xy = TRUE)

# Load North America map
north_america <- ne_countries(continent = "North America", returnclass = "sf")

# Plot the raster with a base map
biasmap<-ggplot() +
  geom_sf(data = north_america, fill = "beige", color = "black") + # Add country borders
  geom_raster(data = bias_df, aes(x = x, y = y, fill=Total_percentage)) + 
  scale_fill_viridis_c(name = "Bias Score") +  # Color scale for better visualization
  coord_sf(xlim = c(-170, -50), ylim = c(10, 80), expand = FALSE) + # Adjust extent
  labs(title = "Spatial Bias in Occurrence Data", x = "Longitude", y = "Latitude") +
  theme_minimal()
```


# Background points
To generate pseudoabsence locations in the area of interest, I will use host occurrence data to define the area from which the presence of TuMV is possible (since it's nested in host occurrences for this model). Alternatively, I could also generate random pseudoabsence locations. Here, I won't. 

NOTE TO SELF: I might want to reconsider the host background occs. Include clean herbarium records for more complete ecological niche of the host. 

### Host occurences
Load presence data for hosts, which will be used to manage sampling bias. There are two options. Either use human observations only or use observations AND herbarium specimens. I think the latter is better...

```{r}
#option 1
#read hesperis matronalis gbif records that are ONLY from recent years and human observations
host.occs<-read.csv("/Users/elizabethlombardi/Dropbox (Old)/Research/Human Preferences Paper (Hesperis)/Hesperis TuMV SDM project/gbif_Hmatronalis_cleanBackgroundOccs.csv")
table(host.occs$basisOfRecord) #these are only human observations. 
min(host.occs$year) #only since 2015

#option 2: read all hesperis matronalis gbif records (includes herbarium specimens)
host.all <- read.delim("~/Dropbox (Old)/Research/Human Preferences Paper (Hesperis)/Hesperis TuMV SDM project/HostGBIF_April242025.csv")
table(host.all$basisOfRecord)
summary(host.all$year)

host.all.cl <- host.all %>%
  filter(!decimalLatitude == "") %>%
  filter(!decimalLongitude== "") %>%
  filter(!year =="") %>%
  filter(!coordinateUncertaintyInMeters > 1000) %>%
  filter(!coordinateUncertaintyInMeters == "")

min(host.all.cl$year)
summary(host.all.cl$coordinateUncertaintyInMeters)
hist(host.all.cl$coordinateUncertaintyInMeters)


#Use GeoThinner package 
library(GeoThinneR)

#thin to 1 kilometer across North America
start.time <- Sys.time()
host.thinned <- thin_points(
  data = host.all.cl,     # Dataframe with coordinates
  lon_col = "decimalLongitude",     # Longitude column name
  lat_col = "decimalLatitude",     # Latitude column name
  method = "distance", # Method for thinning
  thin_dist = 1,      # Thinning distance in km,
  trials = 5,          # Number of trials
  all_trials = FALSE,   # Return all trials
  seed = 123           # Seed for reproducibility
)
end.time <- Sys.time()
time.taken <- round(end.time - start.time,2)
time.taken #1.34 minutes

#explore the thinned dataset
sapply(host.thinned$retained, sum) #retained 13,428 records. Removed 14255 records.

# Transform back to WGS84 if needed
host.thinned <- as_sf(host.thinned)
host.thinned <- st_transform(host.thinned, crs = 4326)


#rename the cleaned full dataset so that code works for either option
host.occs = host.thinned

#make sure there are still lat/long columns
host.occs <- st_transform(host.occs, crs = 4326)
host.occs$decimalLongitude <- sf::st_coordinates(host.occs)[, 1]
host.occs$decimalLatitude <- sf::st_coordinates(host.occs)[, 2]


#compare pre and post thinning: they're the same because of how I downloaded and cleaned, I guess?
dim(host.all.cl)
dim(host.thinned)
dim(host.occs)

```

## Buffer for background sampling
Create a buffer around each known host population (thinned already to 1 kilometer). I first created a 100 kilometer buffer around each location, then 10 kilometers. The larger buffer models (100 kilometer) had better AUC scores for model prediction steps. I'll also try 50 kilometers, which might improve both biological relevance and AUC scores. 
```{r}
bg_buff <- flexsdm::calib_area(
  data = host.occs,
  x = "decimalLongitude",
  y = "decimalLatitude",
  method = c("buffer", width = 50000),  # 0 km buffer; AUC score
  crs = "epsg:4326"
)

rast_buff <- terra::rasterize(bg_buff, rastStack$cropland, field = 0, background = NA) #empty raster with extent and res matching data

#crop to calibration buffer area
env_crop <- terra::crop(rastStack$cropland, bg_buff)
#downsample raster to make it manageable
env_try <- terra::aggregate(env_crop, fact = 4, fun = mean)

gc() #force garbage collection to ensure memory is free for next line
host.occs <- as.data.frame(host.occs)
bg_pnts_buff <- flexsdm::sample_background(data=host.occs,
                                           x="decimalLongitude",
                                           y="decimalLatitude",
                                     calibarea = rast_buff,
                                     rlayer = rast_buff,
                                     n = 5000) #5000 points for now
bg_pnts_buff$pr_ab <- NULL #remove the last column; can I adjust and make this a weight based on host density?
write.csv(bg_pnts_buff, "/Users/elizabethlombardi/Dropbox (Old)/Research/Human Preferences Paper (Hesperis)/Hesperis TuMV SDM project/host_backgroundpnts_hostBuffered.csv", row.names = TRUE)


terra::plot(rastStack$bio1)
terra::plot(bg_buff, add = TRUE)
points(bg_pnts_buff, bg="blue")
mapview(bg_buff)

#get climate data for the background points
bg_clim2 <- terra::extract(rastStack, bg_pnts_buff, ID=FALSE) #this is the host-weighted background bioclim df
bg_clim2 <- na.omit(bg_clim2) #remove simulated background points with NA values (likely water!)

bg2<-cor(bg_clim2, method="spearman")

#dendroram of environmental variables that are correlated
corrdist2 <- 1-abs(bg2)
corrdist2 <- as.dist(corrdist2)
dendro2 <- hclust(corrdist2)
par(mar = c(0, 2.5, 2.5, 0))
plot(dendro2, xlab = "",
     ylab = "1 - absolute correlation",
     sub = "")
abline(h = 0.3,
       col = "red",
       lty = "dotted")

#the correlelogram changes based on the buffer size around hosts! Just a little. Like one or two variables
#based on this correlelogram (whatever), I'll use the [[bracketed]] following predictors in the models:
#1. 4 or 7 [[7]] annual temperature range
#2. 3, 9, 6 or 11 [[11] Mean temp of coldest quarter]
#3. 8 mean temp of wettest quarter
#4. cropland
#5. 1, 5 or 10 [[1]] annual mean temp
#6. popn or footprint [[footprint]]
#7. 2 mean diurnal range
#8. 19, 12, 13, 16 [[12]] annual precipitation
#9. 18 precp of warmest quarter
#10. 15, 14, 17 [[15]] precip seasonality


#Identify which predictors are correlated with each other and reduce colliearity
predsHost.wc <- paste0("bio", c(1, 2, 7, 8, 15, 18, 19)) #start with these based on biology and dendrogram for 50 km
predsHost <- c(predsHost.wc, "cropland", "footprint") 

```

# Model Training
I'm going to use the package enmSdmX for training and prediction. https://github.com/adamlilith/enmSdmX

### prep
```{r}
tumv_clim <- terra::extract(rastStack,
                           occs_vect, 
                           ID=FALSE)

# Vector indicating presence/host background
presbg <- c(rep(1, nrow(occs_vect)), 
             rep(0, nrow(bg_clim2)))
# Combine climate for occurrences and background
clim <- rbind(tumv_clim, bg_clim2) #host background bioclimatic data
train_data <- cbind(presbg, clim) 
train_data #9320 simulated absence points; 207 true presence points (for TuMV)

#scale data for GLM (standardize)
train_data_scaled <- train_data[, c("presbg", predsHost)] #grab only relevant predictors
train_data_scaled[, predsHost] <- scale(train_data_scaled[, predsHost])
train_data_scaled <- train_data_scaled %>%
  mutate(row_id = row_number()) %>%  # Add row numbers to track original rows
  filter(!(row_id %in% (train_data_scaled %>%
                          mutate(row_id = row_number()) %>%
                          dplyr::filter(presbg == 0) %>%
                          slice_sample(n = 2) %>%
                          pull(row_id)))) %>%
  dplyr::select(-row_id)  # Remove temporary row_id column
nrow(train_data_scaled)

centers <- colMeans(train_data[, predsHost]) #remember for later
scales <- apply(train_data[, predsHost], 2, sd)


# Ensure total weights of occurrences is same as total weights of background # sites. Give background points a total weight equal to the sum of the occurrence weights
bg_weights <- rep(sum(na.omit(occ_weights)) / nrow(bg_clim2), nrow(bg_clim2)) 
occ_bg_weights <- c(occ_weights, bg_weights)
anyNA(occ_bg_weights) #2 NA values still; both are TuMV presence rows 45 and 185. From occ_weights
occ_bg_weights <- na.omit(occ_bg_weights) #I have to remove two from 'train_data_scaled', too

#Scaled rasters
scaled_rasts <- terra::scale(rastStack[[predsHost]], center = centers,
                                  scale = scales)

#center training data
train_data_scaled[, predsHost] <- scale(train_data_scaled[, predsHost])

```

###k-folds
```{r}
set.seed(123)  # for reproducibility

folds <- rep(NA, nrow(train_data_scaled))

# Separate presences and background
presFold <- which(train_data_scaled$presbg == 1)
bgFold <- which(train_data_scaled$presbg == 0)

# Assign folds separately for presences and background
folds[presFold] <- sample(rep(1:5, length.out = length(presFold)))
folds[bgFold] <- sample(rep(1:5, length.out = length(bgFold)))

# Add folds as a new column
train_data_scaled$fold <- folds

names(train_data_scaled)

#Create training and testing dataframes
testDF <- train_data_scaled %>%
  dplyr::filter(fold == "1")
trainDF <- train_data_scaled %>%
  dplyr::filter(!fold=="1")

#center testing data
testDF[, predsHost] <- scale(testDF[, predsHost])

#center training data
trainDF[, predsHost] <- scale(trainDF[, predsHost])


```

## SAVE ENVIRONMENT BEFORE MODELING####

```{r}
save.image(file="/Users/elizabethlombardi/Dropbox (Old)/Research/Human Preferences Paper (Hesperis)/Hesperis TuMV SDM project/prepEnvironment.RData")
load("/Users/elizabethlombardi/Dropbox (Old)/Research/Human Preferences Paper (Hesperis)/Hesperis TuMV SDM project/prepEnvironment.RData")
```

### GLM

#### WC+HI
Worldclim plus human impact GLM models. 
```{r}
set.seed(234)

#this step will cycle through all possible models
glm1 <- enmSdmX::trainGLM(data = train_data_scaled, #need to rename variable names earlier in script to get it to run
                                resp = "presbg",
                                preds = predsHost,
                                #w = occ_bg_weights,
                                method = "brglmFit",
                                out = c("model", "tuning"))
summary(glm1$model) 
glm1$tuning #lowest AICc model at top (location 1) == 1 + bio15 + bio7 + footprint 

### Variable contribution?
coefs <- coef(glm1$model)
coefs_no_intercept <- coefs[names(coefs) != "(Intercept)"]
abs_coefs <- abs(coefs_no_intercept)
var_contrib <- 100 * abs_coefs / sum(abs_coefs) #variables in the best model, when glm1 is run with full dataset 
###

# GLM model comparison of training and testing data
glm.predict <- enmSdmX::predictEnmSdm(model = glm1$model, newdata = testDF[,predsHost],
                                         na.rm = TRUE)

roc.glm <- roc(response=testDF$presbg, predictor=glm.predict)
plot(roc.glm)
auc(roc.glm) #0.9517 so this is pretty good. Now it's 0.8651 when I thinned the host dataset. 50 kilometer buffer = 0.946


#create prediction and map it
glm.predict.map <- enmSdmX::predictEnmSdm(model = glm1$model, newdata = scaled_rasts,
                                         na.rm = TRUE)
# Map GLM prediction
terra::plot(glm.predict.map,
            main = "Basic GLM model prediction")
terra::points(occs_vect, pch = 16)


```
#### WC alone
Worldclim data only. No human impact variables (the same WC variables). 

Note that this confirms that the models perform better with human impact variables based on AICc values. 
```{r}
#this step will cycle through all possible models
glm2 <- enmSdmX::trainGLM(data = trainDF, #need to rename variable names earlier in script to get it to run
                                resp = "presbg",
                                preds = predsHost.wc,
                                #w = occ_bg_weights,
                                method = "brglmFit",
                                out = c("model", "tuning"))
summary(glm2$model) 
glm2$tuning #lowest AICc model at top (location 1) == 1 + bio15 + footprint 

# GLM model comparison of training and testing data
glm2.predict <- enmSdmX::predictEnmSdm(model = glm2$model, newdata = testDF[,predsHost],
                                         na.rm = TRUE)

roc.glm2 <- roc(response=testDF$presbg, predictor=glm2.predict)
plot(roc.glm2)
auc(roc.glm2) #0.8099

#create prediction and map it
glm2.predict.map <- enmSdmX::predictEnmSdm(model = glm2$model, newdata = scaled_rasts,
                                         na.rm = TRUE)

plot(glm2.predict.map)


# Map GLM prediction
terra::plot(glm2.predict.map,
            main = "Basic GLM model prediction: WorldClim variables ONLY")
terra::points(occs_vect, pch = 16)


```


### Boosted Regression Tree models
Note that I might want to revise so that all of the environmental predictors are included in BRT modeling. I would need to recreate a new dataset with all of the variables, maybe?
```{r}
set.seed(123)
cores <- 2

#BRT with subset of variables
#dfsub <- train_data_scaled2[1:1049, ] 
brt1 <- enmSdmX::trainBRT(data = train_data_scaled, 
                                resp = "presbg",
                                preds = predsHost,
                                learningRate = 0.001, #conservative learning rate for small sample size
                                treeComplexity = c(2, 3), # interactive effects allowed
                                minTrees = 3000, # minimum trees for reliable model(?), but fast
                                maxTrees = 10000, 
                                tryBy = 'treeComplexity', #could also change to learning rate; roc is slightly worse if using learning rate
                                anyway = TRUE, # return models that did not converge
                                verbose = TRUE,
                                cores = cores)

brt.contributions<-brt1$contributions #for the model on the full dataset
brt1$betas
summary(brt1$contributions$rel.inf) #7.212 is the median relative inf contribution. Maybe keep all with values above that?
print(brt1)


# BRT with only WorldClim predictors
predsHost.wc
brt2 <- enmSdmX::trainBRT(data = trainDF, 
                                resp = "presbg",
                                preds = predsHost.wc,
                                learningRate = 0.001, #conservative learning rate for small sample size
                                treeComplexity = c(2, 3), # interactive effects allowed
                                minTrees = 3000, # minimum trees for reliable model(?), but fast
                                maxTrees = 10000, 
                                tryBy = 'treeComplexity', #could also change to learning rate; roc is slightly worse if using learning rate
                                anyway = TRUE, # return models that did not converge
                                verbose = TRUE,
                                cores = cores)

 brt2$contributions 
summary(brt2$contributions$rel.inf)
print(brt2)


# BRT model compare testing and training data
brt.predict <- enmSdmX::predictEnmSdm(model = brt2, newdata = testDF[,predsHost],
                                         na.rm = TRUE)

roc.brt <- roc(response=testDF$presbg, predictor=brt.predict)
plot(roc.brt)
auc(roc.brt)#0.8669 this includes all variables in the boosted regression tree approach
#0.9532 50 km predsHost (reduced variables). I'm going to reduce the variables!!! USE THIS (brt1)
#0.9731 when I include all variables in BRT model

#BRT prediction mapping
brt.predict.map <- enmSdmX::predictEnmSdm(brt1, scaled_rasts)
plot(brt.predict.map)

```

### Maxent model
Also add a maxent modeling approach. I'll probably take an ensemble model stacking approach to figure it all out later. 

#### MaxNet package
This runs fully in R; there's no need for Java. I also tried the dismo package but the java integration was a pain. 
```{r}
install.packages("maxnet")
library(maxnet)

#prep
env.vars <- train_data_scaled[,-1]
resp.var <-train_data_scaled$presbg

#model
set.seed(123)
maxnet1 <- maxnet::maxnet(p=resp.var, data = env.vars, f=maxnet.formula(resp.var, env.vars))
plot(maxnet1)


#evaluate model
maxnet.probs <- predict(maxnet1, env.vars, type = "cloglog")
plot(maxnet.probs)

# ROC curve and AUC
maxnet.probs <- c(maxnet.probs)
roc.maxnet <- roc(resp.var, maxnet.probs)
plot(roc.maxnet, main = paste("ROC Curve - AUC:", round(auc(roc.maxnet), 3)))
auc(roc.maxnet) #area under the curve is 0.9793


#predict to rasters
# Define a custom prediction wrapper for maxnet
predict_maxnet <- function(model, data) {
  # predict expects a data.frame
  as.numeric(predict(model, newdata = as.data.frame(data), type = "cloglog"))
}

# Now safely predict
maxnet1.pred <- predict(scaled_rasts, maxnet1, clamp=T, type=c("exponential"), na.rm=TRUE)
maxnet2.pred.map <- predict(scaled_rasts, maxnet1, clamp=T, type=c("cloglog"), na.rm=TRUE)

plot(maxnet2.pred)
maxnet2.pred$lyr1

```
This breaks R every time I try to use it. DON'T BOTHER WITH THIS. 


#### enmSdmX 
Alternatively, this package has a maxnet integration that might work without Java...I could try that if I wanted to, but maybe I'll just get the analyses done and figures up for now. 


```{r}
set.seed(123)
cores <- 2

#subdf <- train_data_scaled[1:1000,]
max1 <- enmSdmX::trainMaxNet(
  data = train_data_scaled, 
  resp = "presbg",
  preds = predsHost,
  regMult = 5, 
  jackknife = TRUE,
  verbose = TRUE,
  out = c("models", "tuning"), #with the hash, it returns default of best model (based on AICc)
  cores = cores
  )

coefs <- max1$models[[1]]$betas
abs_coefs <- abs(coefs)
var_contrib <- 100 * abs_coefs / sum(abs_coefs) #relative variable contribution for model with full dataset

# View results
round(var_contrib, 2)

#leave out human impact variables
max2 <- enmSdmX::trainMaxNet(
  data = train_data_scaled, 
  resp = "presbg",
  preds = predsHost.wc,
  regMult = 5, 
  jackknife = TRUE,
  verbose = TRUE,
  #out = c("models", "tuning"), #with the hash, it returns default of best model (based on AICc)
  cores = cores
  )

#Explore the maxent variables
summary(max1$varmax) 
max1$betas
min(max1$tuning$AICc) #2159.634
max1$model$importance




# maxent model comparison of training and testing data
max.predict <- enmSdmX::predictEnmSdm(model = max1, newdata = testDF[,predsHost],
                                         na.rm = TRUE)

roc.max <- roc(response=testDF$presbg, predictor=max.predict)
plot(roc.max)
auc(roc.max) #0.9655 so this is pretty good. Now it's 0.8763. now 0.9515
#0.9464 with 50 km buffer 


#create prediction and map it
max.predict.map <- enmSdmX::predictEnmSdm(model = max1$models[[1]], newdata = scaled_rasts)
plot(max.predict.map)

# Map GLM prediction
terra::plot(max.predict.map,
            main = "Basic Maxnet model prediction")
terra::points(occs_vect, pch = 16)


```
# Mapping 
```{r}
#color palette 
#pal <- cividis(100)
pal <- wes_palette("Zissou1", n=100, type="continuous")
#pal <- colorRampPalette(brewer.pal(11, "PRGn"))

maps <- c(glm.predict.map, brt.predict.map, max.predict.map)

names(maps) <- c('GLM', 'BRT', 'MaxNet')
fun <- function() plot(occs_vect, col='black', pch=16, cex=0, add=TRUE)
plot(maps, col=pal, fun = fun, nc = 4)
```
## Consensus map
Build a consensus map with the three models

```{r}
consensus.map <- c(glm.predict.map, brt.predict.map, max.predict.map)
consensus.map <- mean(consensus.map)

plot(consensus.map, main = "Consensus Prediction Map", col = pal)
```
### GGPLOT version
To match the local scale, add a compass rose, etc. make it a ggplot object and alter aesthetics 

```{r}
# Make sure sf objects are ready
ny_outline <- st_as_sf(ny.sf)
tumv_points <- st_as_sf(tumv.nys.sf)

tumv <- tumv.nys.sf %>%
  mutate(color = ifelse(tumv.binary == 1, "black", "white"))

#PLOT
glmMap1b <- ggplot() +
  geom_raster(data = r_df, aes(x = x, y = y, fill = prediction)) +
  
  scale_fill_gradientn(colors = pal) +
  
  geom_sf(data = ny.sf, fill = NA, color = "black", size = 0.8) +
  
  geom_sf(data = tumv, aes(color = color), size = 2) +
  scale_color_identity() +

   coord_sf() +
  annotation_north_arrow(
    location = "tr",  # top right
    which_north = "true",
    pad_x = unit(0.5, "cm"),
    pad_y = unit(0.5, "cm"),
    style = north_arrow_fancy_orienteering
  ) +
  theme_minimal() +
  labs(title = "Predicted GLM Surface with TUMV Presence/Absence") +
  theme(legend.position = "right")


#save the plot
ggsave("/Users/elizabethlombardi/Dropbox (Old)/Research/Human Preferences Paper (Hesperis)/Hesperis TuMV SDM project/glmMap1b_scaled_withWintTemp.png", width = 8, height = 6, dpi = 300)
```


##Save maps
Create map objects to save 
```{r}
#panel plot
library(patchwork)


```

# Binary prediction maps
Create thresholds

```{r}
model_output <- trainRF(
  data = trainDF,
  resp = ,
  preds = predsHost,
  resp.xy = coordinates(species_data),
  resp.name = "MySpecies"
)


predictions <- predictEnmSdm(
  model = model_output,
  new.env = env_stack
)

threshold <- evalThreshold(
  pres = predictions[species_data$presence == 1],
  contrast = predictions[species_data$presence == 0],
  at = "msss"
)

binary_map <- predictions >= threshold
plot(binary_map, main = "Presence/Absence Map")


  data = trainDF, 
  resp = "presbg",
  preds = predsHost,
  regMult = 5, 
  jackknife = TRUE,
  verbose = TRUE,
  #out = c("models",
  
```

# AUC for models
```{r}
auc(roc.glm) #0.946
auc(roc.brt) #0.9685
auc(roc.max) #0.9655
```



# Write prediction rasters
Fill this in with the model prediction rasters and save them to local computer
```{r}
output_dir <- "/Users/elizabethlombardi/Dropbox (Old)/Research/Human Preferences Paper (Hesperis)/Hesperis TuMV SDM project/Rasters/Model Prediction Rasters/"

# List of rasters to save
rastList.modPredict <- list(
  "glmpredict" = glm.predict.map,
  "brtpredict" = brt.predict.map, 
  "maxnet2" = max.predict.map
)

for (i in names(rastList.modPredict)) {
  writeRaster(
    rastList.modPredict[[i]], 
    filename = file.path(output_dir, paste0(i, ".tif")),
    overwrite = TRUE
  )
}

#just check
raster.files <- list.files(path = "/Users/elizabethlombardi/Dropbox (Old)/Research/Human Preferences Paper (Hesperis)/Hesperis TuMV SDM project/Rasters/Model Prediction Rasters/", pattern = "\\.tif$", full.names = TRUE)
rastStack <- rast(raster.files)


###RELOADING RASTERS
#If they don't have the same number of dimensions when uploading from local computer
rasters <- lapply(raster.files, rast) #read them individually

ref <- rasters[[1]] #bioclim rasters chosen?

# Resample all others to match the reference
rasters_aligned <- lapply(rasters, function(r) {
  resample(r, ref, method = "bilinear")  # use "near" for categorical data
})

rastStack <- rast(rasters_aligned) # will need to rename; code above around line 100

```







